\texttt{thornado}-transport: IMEX schemes for two-moment neutrino transport respecting Fermi-Dirac statistics

### 2018-10-03 Documents/M1/Astronum_2018/manuscript.tex
Update the latest version.

\documentclass[a4paper]{jpconf}
\usepackage{graphicx}
\usepackage{./definitions}
\usepackage{lineno,hyperref}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathrsfs}
\usepackage{color}
\usepackage{url}

@@

\begin{abstract}
We develop implicit-explicit (IMEX) schemes for neutrino transport though a background material in \texttt{thornado} code with two-moment method.
Two-moment method evolves the angular moments of a neutrino phase space distribution function and the moments have upper and lower bounds that introduced by Pauli's exclusion principle.
To ensure the scheme preserves these bounds, a algebraic moment closure based on Fermi-Dirac statistics and a convex-invariant time integrator are demanded.
Severn algebraic closures are compared with these bounds and Cernohorsky \& Bludman closure\cite{cernohorskyBludman_1994} which preserves the bounds is chosen for our IMEX schemes.
For the convex-invariant time integrator, two IMEX scheme named PD-ARSs have been proposed following the work in \cite{chu_etal_2018}.
PD-ARS denotes convex-invariant IMEX Runge-Kutta scheme that has first-order accuracy, high-order accuracy in the streaming limit, and works well in the diffusion limit.
The two PD-ARSs have second-order and third-order accurate explicit strong stability-preserving Runge-Kutta methods\cite{gottlieb_etal_2001} as their explicit part, respectively, and therefore second-order and third-order accurate in the streaming limit.
The accuracy and convex-invariance of PD-ARSs are demonstrated in the numerical tests with a first-order discontinuous Galerkin method for spatial discretization and a simple Lax-Friedrichs flux.

>>>>

\documentclass[a4paper]{jpconf}
\usepackage{graphicx}
\usepackage{./definitions}
\usepackage{lineno,hyperref}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathrsfs}
\usepackage{color}
\usepackage{url}
\usepackage[numbers]{natbib}

@@

\begin{abstract}
We develop implicit-explicit (IMEX) schemes for neutrino transport though a background material in \texttt{thornado} code with two-moment method that evolves the angular moments of a neutrino phase space distribution function.
Considering the upper and lower bounds that introduced by Pauli's exclusion principle on the moments, a algebraic moment closure based on Fermi-Dirac statistics and a convex-invariant time integrator are demanded.
Severn algebraic closures are compared with these bounds and Cernohorsky \& Bludman closure\cite{cernohorskyBludman_1994} which preserves the bounds is chosen for our IMEX schemes.
For the convex-invariant time integrator, two IMEX scheme named PD-ARSs have been proposed following the work in \cite{chu_etal_2018}.
PD-ARS denotes convex-invariant IMEX Runge-Kutta scheme that has first-order accuracy, high-order accuracy in the streaming limit, and works well in the diffusion limit.
Our two PD-ARSs have second-order and third-order explicit strong stability-preserving Runge-Kutta methods\cite{gottlieb_etal_2001} as their explicit part, respectively, and therefore are second-order and third-order accurate in the streaming limit.
The accuracy and convex-invariance of PD-ARSs are demonstrated in the numerical tests with a first-order discontinuous Galerkin(DG) method for spatial discretization and a simple Lax-Friedrichs flux.

### 2018-10-03 Documents/M1/Astronum_2018/sections/Conclusion.tex
Update the latest version.

In the applications motivating this work, the neutrino distribution function can vary from 0 to 1, and we have considered the algebraic closure respecting Fermi-Dirac statistics for both low and high occupied condition.
Different values given by different algebraic closures were compared with the boundary values given by the Fermi-Dirac statistics at different occupation.
Among the algebraic closures we considered, Kershaw\cite{kershaw_1976}, Wilson\cite{wilson_1975,leblanc_1970}, Levermore\cite{levermore_1984}, Minerbo \cite{minerbo_1978}, Janka 1\cite{janka_1991} Janka 2\cite{janka_1992} and Cernohorsky \& Bludman \cite{cernohorskyBludman_1994}, only Cernohorsky \& Bludman closure respecting Fermi-Dirac statistics for any occupancy.
And it is Cernohorsky \& Bludman closure that we employed in the numerical tests in Section~\ref{se:NumericalTests}.

Following the work we have done in \cite{chu_etal_2018}, two PD-ARSs have been proposed.
The one with SSPRK2 is second-order and strong stability-preserving in the streaming limit while other with SSPRK3 is third-order.
Their properties, accuracy and convex-invariant, are demonstrated with numerical tests.

>>>>

In the applications motivating this work, the neutrino distribution function can vary from 0 to 1.
Hence, we have considered the algebraic closure respecting Fermi-Dirac statistics for both low and high occupied condition: different values given by different algebraic closures are compared with the boundary values given by the Fermi-Dirac statistics at different occupation.
Among the seven algebraic closures we considered, Kershaw\cite{kershaw_1976}, Wilson\cite{wilson_1975,leblancWilson_1970}, Levermore\cite{levermore_1984}, Minerbo \cite{minerbo_1978}, Janka 1\cite{janka_1991} Janka 2\cite{janka_1992} and Cernohorsky \& Bludman \cite{cernohorskyBludman_1994}, only Cernohorsky \& Bludman closure respecting Fermi-Dirac statistics for all the occupancy.
So we employ Cernohorsky \& Bludman closure in all the numerical tests in Section~\ref{se:NumericalTests}.

Two PD-ARSs are proposed for the desired IMEX scheme.
The one with SSPRK2 is second-order and strong stability-preserving in the streaming limit while other with SSPRK3 is third-order.
Their properties, both accuracy and convex-invariant, are demonstrated with numerical tests.

### 2018-10-03 Documents/M1/Astronum_2018/sections/SpacialDiscretization.tex
Update the latest version.

\section{Spacial Discretization}\label{se:SpacialDiscretization}

Divide the spatial domain $D$ into $N$ cells and denote the cell as $I_{i}$ with $i = 1,\ldots,N$, so that
\begin{equation*}
D = \cup_{i = 1}^{N} I_{i}.
\end{equation*}
Each of the cell has a volume/length of $\dx$.
Then we have cell average moment:
\begin{equation}
\bcM_{i} = \dfrac{1}{\dx} \int_{I_i}\bcM dx.
\end{equation}
Integrate Eq.~\eqref{eq:momentEquations} for each cell $I_{i}$ to have
\begin{equation}
\dfrac{d \bcM_{i}}{d t} = - \dfrac{1}{\dx} \left( \widehat{\bcF}(\bcM_{i},\bcM_{i+1}) -  \widehat{\bcF}(\bcM_{i-1},\bcM_{i})\right) + \f{1}{\tau}\,\cC(\bcM_{i}),
\end{equation}
where $\widehat{\bcF}(\vect{\cM}_{a},\vect{\cM}_{b})$ is numerical flux and $\f{1}{\tau}\,\cC(\bcM_{i})$ is the integrated collision term.
In this paper we use the simple Lax-Friedrichs flux that given by
\begin{equation}
  \widehat{\bcF}_{LF}(\vect{\cM}_{a},\vect{\cM}_{b})
  =\f{1}{2}\,\big(\,\bcF(\vect{\cM}_{a})+\bcF(\vect{\cM}_{b})-(\,\vect{\cM}_{b}-\vect{\cM}_{a}\,)\,\big).
\end{equation}
Therefore, by treating the flux term explicitly and the collision term implicitly, we have
\begin{align*}
\bcM_{i}^{n+1} = \widetilde{\bcM}^{n}_{i} + \f{\dt}{\tau}\,\cC(\bcM^{n+1}_{i})
\end{align*}
with
\begin{align*}
\widetilde{\bcM}^{n}_{i} 
& = \bcM_{i}^{n} - \dfrac{\dt}{\dx} \left( \widehat{\bcF}(\bcM^{n}_{i},\bcM^{n}_{i+1}) -  \widehat{\bcF}(\bcM^{n}_{i-1},\bcM^{n}_{i})\right)\nonumber \\
& = (1-\beta)\bcM_{i}^{n} + \beta\left[ \f{1}{2}\left( \bcM^{n}_{i+1}-\bcF(\bcM^{n}_{i+1})\right)  + \f{1}{2}\left( \bcM^{n}_{i-1}+\bcF(\bcM^{n}_{i-1})\right)\right]
\end{align*}
where $\beta := \dfrac{\dt}{\dx}$.
Assume $\bcM_{i}^{n}$ is realizable for all $i$. 
Lemma 3 in \cite{chu_2018} says $\bcM^{n+1}_{i}$ remains realizable given $\f{\dt}{\tau} > 0$ and $\widetilde{\bcM}^{n}_{i}$ is realizable.
Substitute the Lax-Friedrichs flux formula in and $\widetilde{\bcM}^{n}_{i}$ becomes a convex combination of $\bcM_{i}^{n}$ and the square bracket term if $\beta \in [0,1]$.
So it is realizable if $\beta \in [0,1]$ and the square bracket term is realizable.
Given by Lemma 2 in \cite{chu_2018}, the square bracket term is realizable.
Therefore, $\bcM^{n+1}_{i}$ remains realizable given $\beta \in [0,1]$.
Note that $\beta \in [0,1]$ is also the CFL condition of the Forward-Euler method employing on a transport only Boltzmann equation.

>>>>

\section{Spacial Discretization}\label{se:SpacialDiscretization}
Here we briefly outline our spacial discretization with the first order discontinuous Galerkin method.
Let the spatial dimension be $d$ and employ Cartesian coordinates.
We divide the spatial domain $D$ evenly into $N$ cells and denote the cell as $\bK_{i}$ with $i = 1,\ldots,N$, so that
\begin{equation*}
D = \cup_{i = 1}^{N} \bK_{i} \quad \text{with} \quad
  \bK_{i}=\{\,\vect{x} : x^{i} \in K^{i} := (\xL^{i},\xH^{i}),~|~i=1,\ldots,d\,\}.
\end{equation*}
Each of the cell has a volume of $|\bK|$
\begin{equation*}
  |\bK| = \int_{\bK}d\vect{x}, \quad\text{where}\quad d\vect{x} = \prod_{i=1}^{d}dx^{i}.  
\end{equation*}
Then we have cell average moment:
\begin{equation}
\bcM_{i} = \dfrac{1}{|\bK|} \int_{\bK_i}\bcM dx.
\end{equation}
Integrate Eq.~\eqref{eq:momentEquations} for each cell $\bK_{i}$ to have
\begin{equation}
\dfrac{d \bcM_{i}}{d t} = - \dfrac{1}{|\bK|} \left( \widehat{\bcF}(\bcM_{i},\bcM_{i+1}) -  \widehat{\bcF}(\bcM_{i-1},\bcM_{i})\right) + \f{1}{\tau}\,\cC(\bcM_{i}),
\end{equation}
where $\widehat{\bcF}(\vect{\cM}_{a},\vect{\cM}_{b})$ is numerical flux and $\f{1}{\tau}\,\cC(\bcM_{i})$ is the integrated collision term.
In this paper we use the simple Lax-Friedrichs flux that given by
\begin{equation}
  \widehat{\bcF}_{LF}(\vect{\cM}_{a},\vect{\cM}_{b})
  =\f{1}{2}\,\big(\,\bcF(\vect{\cM}_{a})+\bcF(\vect{\cM}_{b})-(\,\vect{\cM}_{b}-\vect{\cM}_{a}\,)\,\big).
  \label{eq:Lax-Friedrichs flux}
\end{equation}
Therefore, by treating the flux term explicitly and the collision term implicitly, we have
\begin{align}
\bcM_{i}^{n+1} = \widetilde{\bcM}^{n}_{i} + \f{\dt}{\tau}\,\cC(\bcM^{n+1}_{i})
\label{eq:MomentIMEX}
\end{align}
with
\begin{align}
\widetilde{\bcM}^{n}_{i} 
& = \bcM_{i}^{n} - \dfrac{\dt}{|\bK|} \left( \widehat{\bcF}(\bcM^{n}_{i},\bcM^{n}_{i+1}) -  \widehat{\bcF}(\bcM^{n}_{i-1},\bcM^{n}_{i})\right)\nonumber \\
& = (1-\beta)\bcM_{i}^{n} + \beta\left[ \f{1}{2}\left( \bcM^{n}_{i+1}-\bcF(\bcM^{n}_{i+1})\right)  + \f{1}{2}\left( \bcM^{n}_{i-1}+\bcF(\bcM^{n}_{i-1})\right)\right]
\label{eq:widetildeM}
\end{align}
where $\beta := \frac{\dt}{|\bK|}$.

Assume $\bcM_{i}^{n}$ is realizable for all $i$. 
For Eq.~\eqref{eq:MomentIMEX}, lemma 3 in \cite{chu_etal_2018} says $\bcM^{n+1}_{i}$ remains realizable given $\f{\dt}{\tau} > 0$ and $\widetilde{\bcM}^{n}_{i}$ is realizable.
Substitute the Lax-Friedrichs flux Eq.~\eqref{eq:Lax-Friedrichs flux} in and $\widetilde{\bcM}^{n}_{i}$ becomes Eq.~\eqref{eq:widetildeM}, a convex combination of $\bcM_{i}^{n}$ and the square bracket term if $\beta \in [0,1]$, and it is realizable if $\beta \in [0,1]$ and the square bracket term is realizable.
Given by Lemma 2 in \cite{chu_etal_2018}, the square bracket term is indeed realizable.
Therefore, $\bcM^{n+1}_{i}$ remains realizable given $\beta \in [0,1]$.
Note that $\beta \in [0,1]$ is also the CFL condition of the Forward-Euler method employing on a transport only Boltzmann equation.

### 2018-10-03 Documents/M1/Astronum_2018/sections/TwoMoments.tex
Update the latest version.

The inequalities in Eq.\ref{eq:MomentsBounds} define the realizable $\bcM$ which defines a convex set in $\cJ -\bcH$ space.
As we will see later in Section \ref{se:SpacialDiscretization}, this convexity makes a constraint-preserving spacial discretization possible.

The inequalities in Eq.\ref{eq:eddingtonFactorBounds} desire more attention than have be given.
They have the equal importance as the inequalities in Eq.\ref{eq:MomentsBounds} in preserving the validity of the simulation result.
However, as Fig.~\eqref{fig:EddingtonFactorsWithDifferentClosure} shows, not all the algebraic closures satisfy the Eddington factor bounds Eq.\ref{eq:eddingtonFactorBounds}: Kershaw\cite{kershaw_1976}, Wilson [citation], Levermore\cite{levermore_1984}, Minerbo \cite{minerbo_1978}, Janka 2 [citation] may work fine with neutrino low occupied setting, but not for high occupancy situation; Janka 1 [citation] can be dangerous for both the cases. The only algebraic closure that remains in the bounds among those seven plotted closures is Cernohorsky \& Bludman's \cite{cernohorskyBludman_1994}.
\begin{figure}[h]
  \centering
  \begin{tabular}{cc}
    \includegraphics[width=0.5\textwidth]{figures/Closures0_10}
    \includegraphics[width=0.5\textwidth]{figures/Closures0_90}
  \end{tabular}
   \caption{Plot of Eddington factors $\chi$ versus flux factor $h$ for different values of $\cJ$ for various algebraic closures: $\cJ=0.1$ (left panel, low occupied) and $\cJ=0.9$ (right panel, high occupied).  In each panel we plot the Eddington factors of Kershaw (red), Wilson (yellow), Levermore (green), Minerbo (light blue), Cernohorsky \& Bludman (blue) and Janka (purple and pink) closures.  We also plot $\chi_{\mbox{\tiny min}}$ and $\chi_{\mbox{\tiny max}}$ defined in Eq.~\eqref{eq:eddingtonFactorBounds} (lower and upper dash black lines, respectively).}
  \label{fig:EddingtonFactorsWithDifferentClosure}
\end{figure}

>>>>

The inequalities in Eq.~\eqref{eq:MomentsBounds} define the realizable $\bcM$ which defines a convex set in $\cJ -\bcH$ space.
As we will see later in Section \ref{se:SpacialDiscretization}, this convexity makes a constraint-preserving spacial discretization possible.

The inequalities in Eq.~\eqref{eq:eddingtonFactorBounds} desire more attention than have be given.
They have the equal importance as the inequalities in Eq.~\eqref{eq:MomentsBounds} in preserving the validity of the simulation result.
However, as Fig.~\eqref{fig:EddingtonFactorsWithDifferentClosure} shows, not all the algebraic closures satisfy the Eddington factor bounds Eq.~\eqref{eq:eddingtonFactorBounds}: Kershaw\cite{kershaw_1976}, Wilson\cite{wilson_1975,leblancWilson_1970}, Levermore\cite{levermore_1984}, Minerbo \cite{minerbo_1978}, Janka 2\cite{janka_1992} may work fine with neutrino low occupied setting, but not for high occupancy situation; Janka 1\cite{janka_1991} can be dangerous for both the cases.
The only algebraic closure that remains in the bounds among those seven plotted closures is Cernohorsky \& Bludman's \cite{cernohorskyBludman_1994}.

Though Levermore closure and Minerbo closure don't conserve the bounds with high occupied distribution, they are widely used: {O'Connor} and {Couch}\cite{oConnorCouch_2018}, Pan and et al\cite{pan_etal_2018}, Glas et al\cite{glas_etal_2018} and Just et al\cite{just_etal_2018} use Minerbo closure and Vartanyan et al\cite{vartanyan_etal_2018}, Cabezon et al\cite{cabezon_etal_2018} and Kuroda et al\cite{kuroda_etal_2016} use Levermore clsoure.

\begin{figure}[h]
  \centering
  \begin{tabular}{cc}
    \includegraphics[width=0.5\textwidth]{figures/Closures0_10}
    \includegraphics[width=0.5\textwidth]{figures/Closures0_90}
  \end{tabular}
   \caption{Plot of Eddington factors $\chi$ versus flux factor $h$ for different values of $\cJ$ for various algebraic closures: $\cJ=0.1$ (left panel, low occupied) and $\cJ=0.9$ (right panel, high occupied).  In each panel we plot the Eddington factors of Kershaw (red), Wilson (yellow), Levermore (green), Minerbo (light blue), Cernohorsky \& Bludman (blue) and Janka (purple and pink) closures.  We also plot $\chi_{\mbox{\tiny min}}$ and $\chi_{\mbox{\tiny max}}$ defined in Eq.~\eqref{eq:eddingtonFactorBounds} (lower and upper dash black lines, respectively).}
  \label{fig:EddingtonFactorsWithDifferentClosure}
\end{figure}

### 2018-10-03 Documents/M1/Astronum_2018/sections/intro.tex
Update the latest version.

Nevertheless, solving multi-dimensional Boltzmann transport equation can be expensive.
To balance physical fidelity and computational expediency, an approximation method called two-moment method is established.
In two-moment method, the evolve variables are the zeroth and first angular moments of the distribution function $f$, the spectral particle density $\cJ$ and flux $\bcH$ respectively.

@@

\ee{Some background on current ``state-of-the-art'' simulations and the approach they take.} \cite{vartanyan_etal_2018,oConnorCouch_2018,pan_etal_2018,cabezon_etal_2018,glas_etal_2018,just_etal_2018,kuroda_etal_2016}

To employing IMEX schemes for two-moment neutrino transport, two things need to be chosen carefully: a Fermi-Dirac statistics satisfied algebraic closure for closing the two-moment equations and a convex-invariant diffusion accurate IMEX schemes for ensure a valid result.
Here is the explanation.

First of all, the Boltzmann transport equation is a nonlinear equation.
Its integral form which gives the two-moment method requires the information of $(i+1)$-th order moment for $i$-th order moment evolution.
Therefore, two-moment method needs a algebraic closure which gives a approximated value of the second order moment $\bcK$ given $\cJ$ and $\bcH$.

Second, the angular moments are confined.
Since a neutrino distribution function is confined ($f\in(0,1)$) by the Pauli exclusion principle, its moments as integrals of a confined function over confined domain ($\omega\in\bbS^{2}$) are also confined.
We call the moments satisfying a Fermi-Dirac statistics realizable moments.
Hence, the algebraic closure should give a realizable $\bcK$ and the well-posedness of closure requires realizable $\cJ$ and $\bcH$.

Third, the realizability of $\cJ$ and $\bcH$ on each time step requires a convex-invariant IMEX.
Since the realizable moments form a convex set, it is possible to construct a constraint-preserving IMEX for two-moment neutrino transport method.

Fourth, the physics of neutrino transport in CCSNe requires the IMEX be diffusion limit accurate.
\rc{\cite{gottlieb_etal_2001} showed standard IMEX scheme can't have an order higher than first without a restrict time step requirement.}
To have second-order (or higher-order) accuracy, some correction steps are needed.
However, the correction step can deteriorate the accuracy of the IMEX in the diffusion limit and restrict the time step.
To keep simple, we focus on the IMEX scheme without correction step and require it high-order (second or higher order) in the streaming limit and diffusion accurate.
We call these IMEX scheme PD-ARS.
The study of moment realizability and constraint-preserving diffusion accurate IMEX scheme motivates this work.

>>>>

Nevertheless, solving multi-dimensional Boltzmann transport equation can be expensive.
To balance physical fidelity and computational expediency, an approximation method called two-moment method is established.
In two-moment method, the evolve variables are the zeroth and first angular moments of the distribution function $f$, the spectral particle density $\cJ$ and flux $\bcH$ respectively.
However, the Boltzmann transport equation is a nonlinear equation.
Its integral form which gives the two-moment method requires the information of the second angular moment $\bcK$ to close the system.
Therefore, a algebraic closure which gives a approximated value of $\bcK$ given $\cJ$ and $\bcH$ is needed.
The two-moment method has been widely applied with different algebraic closures, such as the Minerbo\cite{minerbo_1978} closure(e.g.~{O'Connor} and {Couch}\cite{oConnorCouch_2018}, Pan and et al\cite{pan_etal_2018}, Glas et al\cite{glas_etal_2018} and Just et al\cite{just_etal_2018} ) and the Levermore\cite{levermore_1984} closure(e.g.~Vartanyan et al\cite{vartanyan_etal_2018}, Cabezon et al\cite{cabezon_etal_2018} and Kuroda et al\cite{kuroda_etal_2016}).

@@

%\ee{Some background on current ``state-of-the-art'' simulations and the approach they take.} 
To employing IMEX schemes for two-moment neutrino transport, two things need to be chosen carefully: a Fermi-Dirac statistics satisfied algebraic closure for closing the two-moment equations and a convex-invariant diffusion accurate IMEX schemes for ensure a valid result.
Because the angular moments are confined.
Since a neutrino distribution function is confined ($f\in(0,1)$) by the Pauli exclusion principle, its moments as integrals of a confined function over confined domain ($\omega\in\bbS^{2}$) are also confined.
We call the moments satisfying a Fermi-Dirac statistics realizable moments.
Hence, the algebraic closure should give a realizable $\bcK$ and the well-posedness of closure requires realizable $\cJ$ and $\bcH$.
That's why a Fermi-Dirac statistics satisfied algebraic closure is needed.
Then, the realizability of $\cJ$ and $\bcH$ on each time step requires a convex-invariant IMEX.
Since the realizable moments form a convex set, it is possible to construct a constraint-preserving IMEX for two-moment neutrino transport method.
What's more, the physics of neutrino transport in CCSNe requires the IMEX be diffusion limit accurate.

\rc{\cite{gottlieb_etal_2001} showed standard IMEX scheme can't have an order higher than first without a restrict time step requirement.}
To have second-order (or higher-order) accuracy, some correction steps are needed.
Unfortunately, the correction step can deteriorate the accuracy of the IMEX in the diffusion limit and restrict the time step.
To keep simple, we focus on the IMEX scheme without correction step and require it high-order (second or higher order) in the streaming limit and diffusion accurate.
We call these IMEX scheme PD-ARS.
The study of moment realizability and constraint-preserving diffusion accurate IMEX scheme motivates this work.
